---
title: "Differentiated Signal Response Task"
output: 
  html_document: 
    keep_md: no
    toc: true
    toc_float: true
    toc_depth: 4
    collapsed: false
    theme: default
    code_folding: hide
  pdf_document:
    toc: true
    toc_depth: 4
---

## Overview

## Dependencies

Load all required R libraries.

```{r, message=FALSE}
library(ggplot2)  # (Wickham, 2009)
library(dplyr)    # (Wickham et al., 2018)
library(cowplot)  # (Wilke, 2018)
library(viridis)  # (Garnier, 2018)
library(eulerr)
library(tidyr)
```

## Setup

Load data, initial data cleanup, global settings.

```{r}
data_loc <- "../data/mc-reg/max_fit_orgs.csv"
data <- read.csv(data_loc, na.strings="NONE")
data <- subset(data, select = -c(program))

# Specify factors.
data$matchbin_thresh <- factor(data$matchbin_thresh,
                                     levels=c(0, 25, 50, 75))

data$TAG_LEN <- factor(data$TAG_LEN,
                       levels=c(32, 64, 128))
data$DEME_SIZE <- data$DEME_HEIGHT*data$DEME_WIDTH
data$relies_on_reg <- factor(data$relies_on_reg, levels=c(0, 1))
data$relies_on_mem <- factor(data$relies_on_mem, levels=c(0, 1))
data$relies_on_msg <- factor(data$relies_on_msg, levels=c(0, 1))
data$relies_on_imprinting <- factor(data$relies_on_imprinting, levels=c(0, 1))
data$relies_on_repro_tag <- factor(data$relies_on_repro_tag, levels=c(0, 1))
# Define function to summarize regulation/memory configurations.
get_con <- function(reg, mem) {
  if (reg == "0" && mem == "0") {
    return("none")
  } else if (reg == "0" && mem=="1") {
    return("memory")
  } else if (reg=="1" && mem=="0") {
    return("regulation")
  } else if (reg=="1" && mem=="1") {
    return("both")
  } else {
    return("UNKNOWN")
  }
}
# Specify experimental condition for each datum.
data$condition <- mapply(get_con, data$USE_FUNC_REGULATION, data$USE_GLOBAL_MEMORY)
data$condition <- factor(data$condition, levels=c("regulation", "memory", "none", "both"))

data$stochastic <- 1 - data$consistent # Does this program rely on a stochastic strategy?

# Settings for statistical analyses.
alpha <- 0.05
correction_method <- "bonferroni"
theme_set(theme_cowplot())
```

## Multicell Fitness

Here, we give the raw task scores (log scale) for the highest-fitness organism from each run across all environments/conditions. Note that the axes are different for each multicell size.

```{r}
# From: https://stackoverflow.com/questions/23901907/create-a-log-sequence-across-multiple-orders-of-magnitude
lseq <- function(from=1, to=100000, length.out=6) {
  return(exp(seq(log(from), log(to), length.out = length.out)))
}

ggplot(data, aes(x=condition, y=score, color=condition)) +
  geom_boxplot() +
  ggtitle("Scores by Condition") +
  xlab("Condition") +
  ylab("Score (# correct responses)") +
  scale_color_discrete(breaks=c("regulation", "memory", "none", "both"),
                   labels=c("Regulation", "Global Memory", "Neither", "Both")) +
  scale_x_discrete(breaks=c("regulation", "memory", "none", "both"),
                   labels=c("Regulation", "Global\nMemory", "Neither", "Both")) +
  scale_y_continuous(trans="log") +
  theme(legend.position="none") +
  facet_wrap(~DEME_SIZE, scales="free") +
  ggsave("./imgs/mc-reg-scores.png", width=16,height=8)
```

## What do multi-cells rely on to differentiate cell responses?

### Multicell size = 10x10

Maximum of 100 cells per multicell and 10 possible responses.

```{r}
reg_data <- filter(data, condition=="both" & DEME_SIZE==100)

msg_p <- ggplot(reg_data, aes(x=relies_on_msg, fill=relies_on_msg)) +
            geom_bar() +
            geom_text(stat='count', aes(label=..count..), vjust=-.1) +
            ylim(0, 55) +
            ggtitle("Relies on messaging?") +
            theme(legend.position="none",axis.title.x=element_blank())

imp_p <- ggplot(reg_data, aes(x=relies_on_imprinting, fill=relies_on_imprinting)) +
            geom_bar() +
            geom_text(stat='count', aes(label=..count..), vjust=-.1) +
            ylim(0, 55) +
            ggtitle("Relies on imprinting?") +
            theme(legend.position="none",axis.title.x=element_blank())

tag_p <- ggplot(reg_data, aes(x=relies_on_repro_tag, fill=relies_on_repro_tag)) +
            geom_bar() +
            geom_text(stat='count', aes(label=..count..), vjust=-.1) +
            ylim(0, 55) +
            ggtitle("Relies on reproduction tags?") +
            theme(legend.position="none",axis.title.x=element_blank())

reg_p <- ggplot(reg_data, aes(x=relies_on_reg, fill=relies_on_reg)) +
            geom_bar() +
            geom_text(stat='count', aes(label=..count..), vjust=-.1) +
            ylim(0, 55) +
            ggtitle("Relies on regulation?") +
            scale_x_discrete(limits=c("0", "1"), labels=c("0", "1")) +
            theme(legend.position="none",axis.title.x=element_blank())


grid <- plot_grid(reg_p,msg_p,tag_p,imp_p, labels = c('A', 'B', 'C', 'D'), label_size = 12)
print(grid)
save_plot("./imgs/multi-cell-differentiation-strategy-10x10.png", grid, base_height=6, base_width=8)
```

Takeaways:

- We can see that all evolved genotypes from this condition rely on regulation. 
- The majority of genotypes rely on messaging and reproduction tags (i.e., allowing parents to control the 'start' tag for their offspring). 
- About half of the genotypes rely on function regulation imprinting, but a majority do not.

What is the strategy breakdown (in more detail)? 

Each genotype may rely on multiple coordination strategies (i.e., messaging, imprinting, reproduction tags, etc) simultaneously. Here, we use a euler diagram to give a more detailed breakdown of coordination strategies.

```{r}
usage_data <- filter(data, condition=="both" & DEME_SIZE==100)
usage_data <- subset(usage_data, select=c(relies_on_msg,relies_on_imprinting,relies_on_repro_tag,relies_on_mem))
usage_data$msg <- usage_data$relies_on_msg == "1"
usage_data$repro_tag <- usage_data$relies_on_repro_tag == "1"
usage_data$imprinting <- usage_data$relies_on_imprinting == "1"
usage_data$mem <- usage_data$relies_on_mem == "1"
usage_data$none <- !usage_data$msg & !usage_data$repro_tag & !usage_data$imprinting & !usage_data$mem
usage_data <- subset(usage_data, select=c(msg, repro_tag, imprinting, mem, none))
euler(usage_data)
e <- plot(euler(usage_data), legend=TRUE)
save_plot("./imgs/multicell-strategies-breakdown-10x10.png", e, base_width=4, base_height=4)
e
```

Takeaways:

- 1 genotype uses a 'none' strategy. I.e., it uses none of the available _explicit_ mechanisms for coordination. My hypothesis is that this genotype is exploiting the timing of the development phase of this experiment. Cell births that occur later during development have fewer time steps to execute than cells born early during development, allowing multicells to leverage relative execution timing to differentiate.
- The most common combination of mechanisms (18) uses all three of: messaging, reproduction tags, and regulation imprinting for differentiation
- The second-most-common combination of mechanisms (14) uses messaging and reproduction tag control.

What does fitness look like for genotypes relying on different mechanisms?

```{r}
ggplot(reg_data, aes(x=relies_on_repro_tag, y=score, fill=relies_on_repro_tag)) +
  scale_y_continuous(trans="log10") +
  geom_boxplot()

ggplot(reg_data, aes(x=relies_on_imprinting, y=score, fill=relies_on_imprinting)) +
  scale_y_continuous(trans="log10") +
  geom_boxplot()

ggplot(reg_data, aes(x=relies_on_msg, y=score, fill=relies_on_msg)) +
  scale_y_continuous(trans="log10") +
  geom_boxplot()
```

Not sure we can read much into these fitness differentials. A few issues here (nont balanced sets, strategies rely on more than one mechanisms, etc).

### Multicell size = 20x20

Maximum of 400 cells per multicell, 20 possible cellular responses.

```{r}
reg_data <- filter(data, condition=="both" & DEME_SIZE==400)

msg_p <- ggplot(reg_data, aes(x=relies_on_msg, fill=relies_on_msg)) +
            geom_bar() +
            geom_text(stat='count', aes(label=..count..), vjust=-.1) +
            ylim(0, 55) +
            ggtitle("Relies on messaging?") +
            theme(legend.position="none",axis.title.x=element_blank())

imp_p <- ggplot(reg_data, aes(x=relies_on_imprinting, fill=relies_on_imprinting)) +
            geom_bar() +
            geom_text(stat='count', aes(label=..count..), vjust=-.1) +
            ylim(0, 55) +
            ggtitle("Relies on imprinting?") +
            theme(legend.position="none",axis.title.x=element_blank())

tag_p <- ggplot(reg_data, aes(x=relies_on_repro_tag, fill=relies_on_repro_tag)) +
            geom_bar() +
            geom_text(stat='count', aes(label=..count..), vjust=-.1) +
            ylim(0, 55) +
            ggtitle("Relies on reproduction tags?") +
            theme(legend.position="none",axis.title.x=element_blank())

reg_p <- ggplot(reg_data, aes(x=relies_on_reg, fill=relies_on_reg)) +
            geom_bar() +
            geom_text(stat='count', aes(label=..count..), vjust=-.1) +
            ylim(0, 55) +
            ggtitle("Relies on regulation?") +
            scale_x_discrete(limits=c("0", "1"), labels=c("0", "1")) +
            theme(legend.position="none",axis.title.x=element_blank())


grid <- plot_grid(reg_p,msg_p,tag_p,imp_p, labels = c('A', 'B', 'C', 'D'), label_size = 12)
print(grid)
save_plot("./imgs/multi-cell-differentiation-strategy-20x20.png", grid, base_height=6, base_width=8)
```

What is the strategy breakdown (in more detail)? 


```{r}
usage_data <- filter(data, condition=="both" & DEME_SIZE==400)
usage_data <- subset(usage_data, select=c(relies_on_msg,relies_on_imprinting,relies_on_repro_tag,relies_on_mem))
usage_data$msg <- usage_data$relies_on_msg == "1"
usage_data$repro_tag <- usage_data$relies_on_repro_tag == "1"
usage_data$imprinting <- usage_data$relies_on_imprinting == "1"
usage_data$mem <- usage_data$relies_on_mem == "1"
usage_data$none <- !usage_data$msg & !usage_data$repro_tag & !usage_data$imprinting & !usage_data$mem
usage_data <- subset(usage_data, select=c(msg, repro_tag, imprinting, mem, none))
euler(usage_data)
e <- plot(euler(usage_data), legend=TRUE)
save_plot("./imgs/multicell-strategies-breakdown-20x20.png", e, base_width=4, base_height=4)
e
```

What does fitness look like for genotypes relying on different mechanisms?

```{r}
ggplot(reg_data, aes(x=relies_on_repro_tag, y=score, fill=relies_on_repro_tag)) +
  scale_y_continuous(trans="log10") +
  geom_boxplot()

ggplot(reg_data, aes(x=relies_on_imprinting, y=score, fill=relies_on_imprinting)) +
  scale_y_continuous(trans="log10") +
  geom_boxplot()

ggplot(reg_data, aes(x=relies_on_msg, y=score, fill=relies_on_msg)) +
  scale_y_continuous(trans="log10") +
  geom_boxplot()
```





