---
title: "Demonstrating Genetic Regulation in SignalGP - Analysis - Differentiated Signal Response Task"
output: 
  html_document: 
    keep_md: no
    toc: true
    toc_float: true
    toc_depth: 4
    collapsed: false
    theme: default
    code_folding: hide
  pdf_document:
    toc: true
    toc_depth: 4
---

## Overview

## Dependencies

Load all required R libraries.

```{r, message=FALSE}
library(ggplot2)  # (Wickham, 2009)
library(dplyr)    # (Wickham et al., 2018)
library(cowplot)  # (Wilke, 2018)
library(viridis)  # (Garnier, 2018)
library(eulerr)
library(tidyr)
```

## Setup

Load data, initial data cleanup, global settings.

```{r}
data_loc <- "../data/mc-reg/max_fit_1500.csv"
data <- read.csv(data_loc, na.strings="NONE")
data <- subset(data, select = -c(program))

# Specify factors.
data$matchbin_thresh <- factor(data$matchbin_thresh,
                                     levels=c(0, 25, 50, 75))

data$TAG_LEN <- factor(data$TAG_LEN,
                       levels=c(32, 64, 128))

data$relies_on_reg <- factor(data$relies_on_reg, levels=c(0, 1))
data$relies_on_mem <- factor(data$relies_on_mem, levels=c(0, 1))
data$relies_on_msg <- factor(data$relies_on_msg, levels=c(0, 1))
data$relies_on_imprinting <- factor(data$relies_on_imprinting, levels=c(0, 1))
data$relies_on_repro_tag <- factor(data$relies_on_repro_tag, levels=c(0, 1))
# Define function to summarize regulation/memory configurations.
get_con <- function(reg, mem) {
  if (reg == "0" && mem == "0") {
    return("none")
  } else if (reg == "0" && mem=="1") {
    return("memory")
  } else if (reg=="1" && mem=="0") {
    return("regulation")
  } else if (reg=="1" && mem=="1") {
    return("both")
  } else {
    return("UNKNOWN")
  }
}
# Specify experimental condition for each datum.
data$condition <- mapply(get_con, data$USE_FUNC_REGULATION, data$USE_GLOBAL_MEMORY)
data$condition <- factor(data$condition, levels=c("regulation", "memory", "none", "both"))

data$stochastic <- 1 - data$consistent # Does this program rely on a stochastic strategy?

# Settings for statistical analyses.
alpha <- 0.05
correction_method <- "bonferroni"
```

## fitness

### scores by condition

Here, we give the raw task scores for the highest-fitness organism from each run across all environments/conditions.

```{r}
# From: https://stackoverflow.com/questions/23901907/create-a-log-sequence-across-multiple-orders-of-magnitude
lseq <- function(from=1, to=100000, length.out=6) {
  return(exp(seq(log(from), log(to), length.out = length.out)))
}

ggplot(data, aes(x=condition, y=score, color=condition)) +
  geom_boxplot() +
  ggtitle("Scores by Condition") +
  xlab("Condition") +
  ylab("Score (# correct responses)") +
  scale_color_discrete(breaks=c("regulation", "memory", "none", "both"),
                   labels=c("Regulation", "Global Memory", "Neither", "Both")) +
  scale_x_discrete(breaks=c("regulation", "memory", "none", "both"),
                   labels=c("Regulation", "Global\nMemory", "Neither", "Both")) +
  scale_y_continuous(trans="log", limits=c(1, 1e+20), breaks=c(lseq(1, 1e+20, 5))) +
  theme(legend.position="none") +
  ggsave("./imgs/mc-reg-scores.png", width=16,height=8)
```

### What do multi-cells rely on to differentiate cell responses?

```{r}
reg_data <- filter(data, condition=="both")

msg_p <- ggplot(reg_data, aes(x=relies_on_msg, fill=relies_on_msg)) +
            geom_bar() +
            ylim(0, 50) +
            ggtitle("Relies on messaging?") +
            theme(legend.position="none",axis.title.x=element_blank())

imp_p <- ggplot(reg_data, aes(x=relies_on_imprinting, fill=relies_on_imprinting)) +
            geom_bar() +
            ylim(0, 50) +
            ggtitle("Relies on imprinting?") +
            theme(legend.position="none",axis.title.x=element_blank())

tag_p <- ggplot(reg_data, aes(x=relies_on_repro_tag, fill=relies_on_repro_tag)) +
            geom_bar() +
            ylim(0, 50) +
            ggtitle("Relies on reproduction tags?") +
            theme(legend.position="none",axis.title.x=element_blank())

reg_p <- ggplot(reg_data, aes(x=relies_on_reg, fill=relies_on_reg)) +
            geom_bar() +
            ylim(0, 50) +
            ggtitle("Relies on regulation?") +
            scale_x_discrete(limits=c("0", "1"), labels=c("0", "1")) +
            theme(legend.position="none",axis.title.x=element_blank())


grid <- plot_grid(reg_p,msg_p,tag_p,imp_p, labels = c('A', 'B', 'C', 'D'), label_size = 12)
print(grid)
save_plot("./imgs/multi-cell-differentiation-strategy.png", grid, base_height=6, base_width=8)
```

What are the fitnesses of demes that relies on each strategy?

```{r}
ggplot(reg_data, aes(x=relies_on_repro_tag, y=score, fill=relies_on_repro_tag)) +
  scale_y_continuous(trans="log10") +
  geom_boxplot()

ggplot(reg_data, aes(x=relies_on_imprinting, y=score, fill=relies_on_imprinting)) +
  scale_y_continuous(trans="log10") +
  geom_boxplot()

ggplot(reg_data, aes(x=relies_on_msg, y=score, fill=relies_on_msg)) +
  scale_y_continuous(trans="log10") +
  geom_boxplot()
```

What is the strategy breakdown (in more detail)?

```{r}
usage_data <- filter(data, condition=="both")
usage_data <- subset(usage_data, select=c(relies_on_msg,relies_on_imprinting,relies_on_repro_tag))
usage_data$msg <- usage_data$relies_on_msg == "1"
usage_data$repro_tag <- usage_data$relies_on_repro_tag == "1"
usage_data$imprinting <- usage_data$relies_on_imprinting == "1"
usage_data$none <- !usage_data$msg & !usage_data$repro_tag & !usage_data$imprinting
usage_data <- subset(usage_data, select=c(msg, repro_tag, imprinting, none))
euler(usage_data)
# png("./plot.png", width=4, height=4)
e <- plot(euler(usage_data), legend=TRUE)
save_plot("./imgs/multicell-strategies-breakdown.png", e, base_width=4, base_height=4)
# dev.off()
```


