---
title: "Demonstrating Genetic Regulation in SignalGP - Analysis - Repeated Signal Task"
output: 
  html_document: 
    keep_md: no
    toc: true
    toc_float: true
    toc_depth: 4
    collapsed: false
    theme: default
    code_folding: hide
  pdf_document:
    toc: true
    toc_depth: 4
---

## Overview

The repeated signal task requires organisms to coordinate their behavior with a changing environment. 
In the repeated signal task, organisms are evaluated in an environment that cycles through $K$ independent states (always in the same order), each lasting a fixed number of time steps. 
To maximize fitness, an organism must express the appropriate response in to each environmental state. 
The capacity to exhibit different responses to a particular signal requires organisms to regulate their response to the signal.

In classic SignalGP, organisms must use their global memory buffer to track the environment and respond to the environmental signals appropriately.
In regulation-enabled SignalGP, organisms can leverage global memory (i.e., memory shared among a program's threads that is persistent between environments; all local thread memory does not persist between environments) or regulation to shift their responses to the environment signal over time.

This document is a companion to our submission to the 2020 Artificial Life conference. Here, we provide our analyses of our repeated signal task results.

We evolved populations of 1,000 SignalGP organisms to solve the repeated signal task at four complexity levels: $K=2$, $K=4$, $K=8$, $K=16$ (where $K$ denotes the number of environmental states). $K=16$ is the most complex (and thus challenging) environment, and $K=2$ is the simplest (and thus least challenging) environment.
We evolved populations for 10,000 generations, or until an organism capable of perfectly solving the task evolved (able to correctly respond in all environments).

For each environment complexity level, we ran four experimental conditions:

1. Classic SignalGP where programs have access to a global memory buffer but not to function regulation
2. Regulation-augmented SignalGP where programs have access to both global memory and to function regulation
3. A regulation-only control SignalGP where programs have access to function regulation but not a global memory buffer
4. A control version of SignalGP where programs have access to neigher function regulation nor a global memory buffer (this version of SignalGP should be incapable of solving the repeated signal task)

We ran 50 replicates of each condition.

## Dependencies

Load all required R libraries.

```{r, message=FALSE}
library(ggplot2)  # (Wickham, 2009)
library(tidyr)    # (Wickham and Henry, 2018)
library(dplyr)    # (Wickham et al., 2018)
library(cowplot)  # (Wilke, 2018)
library(viridis)  # (Garnier, 2018)
library(reshape2)
library(RColorBrewer)
library(igraph)
library(ggplotify)
```

## Setup

Load data, initial data cleanup, global settings.

```{r}
data_dir <- "/Users/amlalejini/devo_ws/signalgp-genetic-regulation/experiments/alife-2020-sgp-reg/data/alt-sig/"
max_fit_org_data_loc <- paste(data_dir, "max_fit_orgs.csv", sep="")
max_fit_org_data <- read.csv(max_fit_org_data_loc, na.strings="NONE")
max_fit_org_data <- subset(max_fit_org_data, select = -c(program))

# Specify factors.
max_fit_org_data$matchbin_thresh <- factor(max_fit_org_data$matchbin_thresh,
                                           levels=c(0, 25, 50, 75))
max_fit_org_data$NUM_SIGNAL_RESPONSES <- factor(max_fit_org_data$NUM_SIGNAL_RESPONSES,
                                                levels=c(2, 4, 8, 16, 32))
max_fit_org_data$NUM_ENV_CYCLES <- factor(max_fit_org_data$NUM_ENV_CYCLES,
                                          levels=c(2, 4, 8, 16, 32))
max_fit_org_data$TAG_LEN <- factor(max_fit_org_data$TAG_LEN,
                                   levels=c(32, 64, 128))
# Define function to summarize regulation/memory configurations.
get_con <- function(reg, mem) {
  if (reg == "0" && mem == "0") {
    return("none")
  } else if (reg == "0" && mem=="1") {
    return("memory")
  } else if (reg=="1" && mem=="0") {
    return("regulation")
  } else if (reg=="1" && mem=="1") {
    return("both")
  } else {
    return("UNKNOWN")
  }
}
# Specify experimental condition for each datum.
max_fit_org_data$condition <- mapply(get_con, 
                                     max_fit_org_data$USE_FUNC_REGULATION,
                                     max_fit_org_data$USE_GLOBAL_MEMORY)
max_fit_org_data$condition <- factor(max_fit_org_data$condition, 
                                     levels=c("regulation", "memory", "none", "both"))

max_fit_org_data$stochastic <- 1 - max_fit_org_data$consistent # Does this program rely on a stochastic strategy?

# Settings for statistical analyses.
alpha <- 0.05
correction_method <- "bonferroni"
theme_set(theme_cowplot()) # Configure our default graphing theme
```

## Does regulation-enabled SignalGP solve the repeated signal task?

We expected populations with access to function regulation to be most successful on the repeated signal task. 
Indeed, we see that conditions where function regulation is enabled yield the most fit organisms (and the most solutions).

### Number of evolved solutions per condition

An organism is categorized as a 'solution' if it can correctly respond in each of the $K$ environment states.

```{r, echo=TRUE}

label_lu <- c(
  "2" = "2 Environments", 
  "4" = "4 Environments", 
  "8" = "8 Environments", 
  "16" = "16 Environments", 
  "32" ="32 Environments"
)


sol_data <- filter(max_fit_org_data, solution=="1")
ggplot(sol_data, aes(x=condition, fill=condition)) +
  geom_bar() +
  geom_text(stat="count", aes(label=..count..), position=position_dodge(0.9), vjust=0) +
  ggtitle("Solution Counts by Condition and Environment Complexity") +
  xlab("Condition") +
  scale_y_continuous(name="# evolved solutions", breaks=seq(0, 50, 10), limits=c(0, 52)) +
  scale_fill_discrete(name="Condition:",
                      limits=c("regulation", "memory", "both", "none"),
                      labels=c("Regulation-only", "Memory-only", "Both", "Neither")) +
  scale_x_discrete(limits=c("regulation", "memory", "both", "none"),
                   labels=c("Regulation", "Memory", "Both", "Neither")) +
  facet_wrap(~ NUM_SIGNAL_RESPONSES, nrow=1, labeller=labeller(NUM_SIGNAL_RESPONSES=label_lu)) +
  theme(legend.position="bottom", axis.text.x=element_blank(), axis.title.x=element_blank()) +
  ggsave("./imgs/repeated-signal-solultion-cnts.png", width=8,height=2.75)
```

Note that in the lowest-complexity environment ($K=2$), two runs using classic SignalGP digital organisms produce solutions. This, in addition to hand-coded programs, confirms that it is _possible_ to solve the repeated signal task without regulation. However, these data seem to indicate that using the global memory buffer as it is currently designed/implemented is more challenging than using function regulation.

We use a Fisher's exact test to determine if there are significant differences (p < `r alpha`) between the numbers of solutions found in the regulation-augmented SignalGP condition and all other conditions (within each environment complexity level). 
We correct for multiple comparisons using the `r correction_method` method.
Indeed, our analyses (given below), confirm our visual intuitions: conditions with access to function regulation outperform those without access to function regulation.

```{r, echo=TRUE, results="asis"}
environments <- c("2", "4", "8", "16")

for (env in environments) {
  
  env_data <- filter(max_fit_org_data, NUM_SIGNAL_RESPONSES==env)
  cat("#### Environment ", env, " - statistical analysis of solution counts  \n")
  # Extract successes/fails for each condition.
  mem_success_cnt <- nrow(filter(env_data, solution=="1" & condition=="memory"))
  mem_fail_cnt <- nrow(filter(env_data, condition=="memory")) - mem_success_cnt
  reg_success_cnt <- nrow(filter(env_data, solution=="1" & condition=="regulation"))
  reg_fail_cnt <- nrow(filter(env_data, condition=="regulation")) - reg_success_cnt
  both_success_cnt <- nrow(filter(env_data, solution=="1" & condition=="both"))
  both_fail_cnt <- nrow(filter(env_data, condition=="both")) - both_success_cnt
  none_success_cnt <- nrow(filter(env_data, solution=="1" & condition=="none"))
  none_fail_cnt <- nrow(filter(env_data, condition=="none")) - none_success_cnt
  
  # Regulation-enabled SGP vs. Classic SGP
  mem_sgp_table <- matrix(c(both_success_cnt, 
                            mem_success_cnt, 
                            both_fail_cnt, 
                            mem_fail_cnt), nrow=2)
  rownames(mem_sgp_table) <- c("both", "mem-only")
  colnames(mem_sgp_table) <- c("success", "fail")
  mem_sgp_fishers <- fisher.test(mem_sgp_table)
  
  # Regulation-enabled SGP vs. Regulation-only SGP
  reg_sgp_table <- matrix(c(both_success_cnt, 
                            reg_success_cnt, 
                            both_fail_cnt, 
                            reg_fail_cnt), nrow=2)
  rownames(reg_sgp_table) <- c("both", "reg-only")
  colnames(reg_sgp_table) <- c("success", "fail")
  reg_sgp_fishers <- fisher.test(reg_sgp_table)
  
  # Regulation-enabled SGP vs. No reg/mem SGP
  none_sgp_table <- matrix(c(both_success_cnt, 
                             none_success_cnt, 
                             both_fail_cnt, 
                             none_fail_cnt), nrow=2)
  rownames(none_sgp_table) <- c("both", "none")
  colnames(none_sgp_table) <- c("success", "fail")
  none_sgp_fishers <- fisher.test(none_sgp_table)
  
  # Adjust for multiple comparisons
  adjusted <- p.adjust(p=c(mem_sgp_fishers$p.value, 
                           reg_sgp_fishers$p.value, 
                           none_sgp_fishers$p.value), method=correction_method)
  mem_sgp_fishers$p.adjusted <- adjusted[1]
  reg_sgp_fishers$p.adjusted <- adjusted[2]
  none_sgp_fishers$p.adjusted <- adjusted[3]
  
  # Summarize!
  no_sig_comps <- TRUE
  cat("Significant comparisons (Fisher's exact test, p < ", alpha, ",",
      correction_method, " correction for multiple comparisons):  \n\n")
  if (mem_sgp_fishers$p.adjusted < alpha) {
    cat("- Regulation-augmented SignalGP (regulation & global memory) vs. Classic SignalGP (global memory only )  \n")
    no_sig_comps <- FALSE
  }
  if (reg_sgp_fishers$p.adjusted < alpha) {
    cat("- Regulation-augmented SignalGP (regulation & global memory) vs. Regulation-only SignalGP (global memory disabled)  \n")
    no_sig_comps <- FALSE
  }
  if (none_sgp_fishers$p.adjusted < alpha) {
    cat("- Regulation-augmented SignalGP (regulation & global memory) vs. Regulation- and Memory-Disabled SignalGP  \n")
    no_sig_comps <- FALSE
  }
  if (no_sig_comps) {
    cat("- NONE  \n")
  }
  
  cat("\n")
  cat("**Results - Regulation-augmented SignalGP (regulation & global memory) vs. Classic SignalGP (global memory only)**  \n")
  cat("```\n")
  print(mem_sgp_table)
  print(mem_sgp_fishers)
  cat("p.adjusted = ", mem_sgp_fishers$p.adjusted, " (method = ", correction_method, ")", "  \n")
  cat("```\n")
  
  cat("**Results - Regulation-augmented SignalGP (regulation & global memory) vs. Regulation-only SignalGP (global memory disabled)**  \n")
  cat("```\n")
  print(reg_sgp_table)
  print(reg_sgp_fishers)
  cat("p.adjusted = ", reg_sgp_fishers$p.adjusted, " (method = ", correction_method, ")", "  \n")
  cat("```\n")
  
  cat("**Results - Regulation-augmented SignalGP (regulation & global memory) vs. Regulation- and Memory-Disabled SignalGP**  \n")
  cat("```\n")
  print(none_sgp_table)
  print(none_sgp_fishers)
  cat("p.adjusted = ", none_sgp_fishers$p.adjusted, " (method = ", correction_method,")", "  \n")
  cat("```\n")
  cat("\n")

}
```

### Repeated signal task scores by condition

Here, we give the raw task scores for the highest-fitness organism from each run across all environments/conditions.

```{r}
ggplot(max_fit_org_data, aes(x=condition, y=score, color=condition)) +
  geom_boxplot() +
  ggtitle("Scores by Condition and Environment Complexity") +
  xlab("Condition") +
  ylab("Score (# correct responses)") +
  scale_color_discrete(breaks=c("regulation", "memory", "none", "both"),
                       labels=c("Regulation", "Global Memory", "Neither", "Both")) +
  scale_x_discrete(breaks=c("regulation", "memory", "none", "both"),
                   labels=c("R", "M", "N", "R & M")) +
  facet_wrap(~ NUM_SIGNAL_RESPONSES, 
             scales="free_y", 
             labeller=labeller(NUM_SIGNAL_RESPONSES=label_lu)) +
  ggsave("./imgs/repeated-signal-scores.png", width=16,height=8)
```

### How long did it take for solutions to arise?

Do some conditions lead to the evolution of solutions faster than other conditions? Here, we look at the generation at which solutions arise. Note that it is not a truly fair comparison to make between conditions in which not an equal number of replicates produce solutions. 

If we wanted to make strong claims based on the time it took for solutions to evolve, we should pick an arbitrary proportion, e.g., 50%. Then, re-run a number of replicates for each condition for as long as it takes 50% of them to produce a solution. We could then compare the generations at which solutions evolved in the first 50% of the replicates to produce a solution.

TL;DR: take the next few analyses with a grain of salt. These are primarily exploratory.

```{r}
ggplot(filter(max_fit_org_data, solution=="1"),
       aes(x=condition, y=update, color=condition)) +
  geom_boxplot() +
  ggtitle("Time to solution") +
  ylab("Generation first solution evolved (log scale)") +
  scale_color_discrete(name="Condition: ",
                       breaks=c("regulation", "memory", "none", "both"),
                       labels=c("Regulation", "Global Memory", "Neither", "Both")) +
  scale_x_discrete(breaks=c("regulation", "memory", "none", "both"),
                   labels=c("Regulation", "Global\nMemory", "Neither", "Both")) +
  scale_y_continuous(limits=c(0, 10000), 
                     breaks=c(0, 10, 100, 1000, 10000), 
                     trans="pseudo_log") +
  facet_wrap(~ NUM_SIGNAL_RESPONSES, 
             nrow=1, 
             labeller=labeller(NUM_SIGNAL_RESPONSES=label_lu)) +
  theme(legend.position="bottom", 
        axis.text.x=element_blank(), 
        axis.title.x=element_blank()) +
  ggsave("./imgs/repeated-signal-solve-time.png",
         width=8,
         height=4)
```

#### 2-state environment

```{r}
env_2_sol_data <- filter(max_fit_org_data, 
                         solution=="1" & NUM_SIGNAL_RESPONSES==2)
# First, kruskal-wallis to test for significant differences among groups
env2_kt <- kruskal.test(update ~ condition,
                        data=env_2_sol_data)
print(env2_kt)
# Next, if kruskal-wallis is significant, perform a pairwise wilcoxon rank-sum test
if (env2_kt$p.value < alpha) {
  # x: response vector; g: grouping vector
  env2_wt <- pairwise.wilcox.test(x=env_2_sol_data$update, 
                                  g=env_2_sol_data$condition, 
                                  p.adjust.method=correction_method, 
                                  exact=FALSE)
  print(env2_wt)
} else {
  print("No significant difference.")
}
```

Here, the solutions from regulation-enabled conditions arose sooner than solutions from memory-only conditions.

#### 4-state environment

```{r}
env_4_sol_data <- filter(max_fit_org_data, 
                         solution=="1" & NUM_SIGNAL_RESPONSES==4)
# First, kruskal-wallis to test for significant differences among groups
env4_kt <- kruskal.test(update ~ condition, 
                        data=env_4_sol_data)
print(env4_kt)
# Next, if kruskal-wallis is significant, perform a pairwise wilcoxon rank-sum test
if (env4_kt$p.value < alpha) {
  # x: response vector; g: grouping vector
  env4_wt <- pairwise.wilcox.test(x=env_4_sol_data$update, 
                                  g=env_4_sol_data$condition, 
                                  p.adjust.method=correction_method, 
                                  exact=FALSE)
  print(env4_wt)
} else {
  print("No significant difference.")
}
```

Here, only regulation-enabled treatments yielded solutions, and there is no significant difference between the two regulation-enabled conditions.

#### 8-state environment

```{r}
env_8_sol_data <- filter(max_fit_org_data,
                         solution=="1" & NUM_SIGNAL_RESPONSES==8)
# First, kruskal-wallis to test for significant differences among groups
env8_kt <- kruskal.test(update ~ condition, 
                        data=env_8_sol_data)
print(env8_kt)
# Next, if kruskal-wallis is significant, perform a pairwise wilcoxon rank-sum test
if (env8_kt$p.value < alpha) {
  # x: response vector; g: grouping vector
  env8_wt <- pairwise.wilcox.test(x=env_8_sol_data$update, 
                                  g=env_8_sol_data$condition, 
                                  p.adjust.method=correction_method, 
                                  exact=FALSE)
  print(env8_wt)
} else {
  print("No significant difference.")
}
```

Here, only regulation-enabled treatments yielded solutions, and there is no significant difference between the two regulation-enabled conditions.

#### 16-state environment

```{r, echo=TRUE}
# First, kruskal-wallis to test for significant differences among groups
env_16_sol_data <- filter(max_fit_org_data, 
                          solution=="1" & NUM_SIGNAL_RESPONSES==16)
env16_kt <- kruskal.test(update ~ condition, 
                         data=env_16_sol_data)
print(env16_kt)
# Next, if kruskal-wallis is significant, perform a pairwise wilcoxon rank-sum test
if (env16_kt$p.value < alpha) {
  # x: response vector; g: grouping vector
  env16_wt <- pairwise.wilcox.test(x=env_16_sol_data$update, 
                                   g=env_16_sol_data$condition, 
                                   p.adjust.method=correction_method, 
                                   exact=FALSE)
  print(env16_wt)
} else {
  print("No significant difference.")
}
```

Here, only regulation-enabled treatments yielded solutions, and there is no significant difference between the two regulation-enabled conditions.

Again, these comparisons are exploratory. Because of the point we discuss at the beginning of this timing section, it isn't necessarily fair to draw solid conclusions from comparing these 'time to solution' data.

## Do solutions rely on function regulation or global memory access for signal-response plasticity?

Here, we take a closer at the strategies employed by solutions evolved across environment complexities.
For each evolved solution, we independently knocked out (disabled) function regulation and global memory access. We measured the fitness effects of knocking each of function regulation and global memory, and if a knockout resulted in a decrease in fitness, we labeled that organism as relying on that functionality (global memory or function regulation) for success.

The graph below gives the proportion of solutions that rely on just regulation, just global memory, and both global memory and regulation for each of the regulation-only, global-memory-only (classic SignalGP), and regulation-and-memory (regulation-augmented SignalGP) conditions.
We can see that in conditions where organisms have access to either just regulation or both regulation and global memory, evolved solutions rely _only_ on regulation. In conditions where memory is the only mechanism for solving the repeated signal task, we see that all evolved solutions (only 2) rely only on memory.

These data show that, in SignalGP's current configuration of function regulation and global memory, evolution 'prefers' function regulation for solving the repeated signal task. 


```{r}
reg_and_mem_solutions <- filter(max_fit_org_data, 
                                solution=="1")
get_strategy <- function(use_reg, use_mem) {
  if (use_reg=="0" && use_mem=="0") {
    return("use neither")
  } else if (use_reg=="0" && use_mem=="1") {
    return("use memory")
  } else if (use_reg=="1" && use_mem=="0") {
    return("use regulation")
  } else if (use_reg=="1" && use_mem=="1") {
    return("use both")
  } else {
    return("UNKNOWN")
  }
}
# Specify experimental conditions (to make labeling easier).
reg_and_mem_solutions$strategy <- mapply(get_strategy,
                                         reg_and_mem_solutions$relies_on_regulation,
                                         reg_and_mem_solutions$relies_on_global_memory)
reg_and_mem_solutions$strategy <- factor(reg_and_mem_solutions$strategy, 
                                         levels=c("use regulation", 
                                                  "use memory", 
                                                  "use neither", 
                                                  "use both"))

ggplot(reg_and_mem_solutions, 
       mapping=aes(x=NUM_SIGNAL_RESPONSES, fill=strategy)) +
  geom_bar(position="fill", stat="count") +
  geom_text(stat='count', 
            mapping=aes(label=..count..), 
            position=position_fill(vjust=-0.03)) +
  ylab("% of Solutions") +
  xlab("Environment Complexity") +
  scale_fill_discrete(name="Strategy:", 
                      breaks=c("use regulation", 
                               "use memory", 
                               "use neither", 
                               "use both"),
                      labels=c("Use regulation (only)", 
                               "Use global memory (only)", 
                               "Use neither", 
                               "Use both")) +
  facet_wrap(~condition) +
  theme(legend.position = "bottom") +
  ggsave("./imgs/repeated-signal-strategies.png",
         width=8,
         height=4)
```

### Are evolved programs relying on stochastic strategies?

To confirm that evolved organisms were not relying on stochastic approaches to solve the repeated signal task, we tested each of the highest-fitness individuals in the population at the end of each run three times. If an organism's phenotype was not identical across each of the three trials, we labeled is as using a stochastic strategy.
From the graph below, we see no evidence of evolved organisms relying on stochastic strategies to solve the repeated signal task. I.e., all organisms responded consistently across trials.
Note, this is unsurprising, as we do not give programs access to instructions capable of generating random values.

```{r}
max_fit_org_data$stochastic <- factor(max_fit_org_data$stochastic,
                                      levels=c(0, 1))
ggplot(max_fit_org_data, aes(x=condition, fill=stochastic)) +
  geom_bar(stat="count") +
  ggtitle("Stochastic Strategies?") +
  ylab("# Replicates") +
  ylim(0, 50) +
  scale_fill_discrete(name="Strategy",
                       limits=c(0, 1),
                       labels=c("Deterministic", "Stochastic")) +
  scale_x_discrete(name="Condition",
                   breaks=c("regulation", "memory", "none", "both"),
                   labels=c("Regulation", "Global\nMemory", "Neither", "Both")) +
  facet_wrap(~ NUM_SIGNAL_RESPONSES, 
             labeller=labeller(NUM_SIGNAL_RESPONSES=label_lu))
```


### What types of regulation are solutions relying on?

Do solutions that rely on regulation primarily use up-regulation or down-regulation?

Note that we're only looking at solutions evolved under the 'both' condition.

This first graph is based on the number of times a function call state results in up-regulation versus down-regulation.

```{r}
melted <- melt(filter(max_fit_org_data, 
                      condition=="both" & solution=="1"), 
               variable.name="call_reg_type",
               value.name="call_reg_cnt",
               measure.vars=c("call_promoted_cnt", "call_repressed_cnt"))
ggplot(melted, aes(y=call_reg_cnt, x=NUM_SIGNAL_RESPONSES, color=call_reg_type)) +
  geom_boxplot() +
  ylab("Regulation Calls") +
  xlab("Environment States") +
  scale_color_discrete(name="Type:",
                       limits=c("call_promoted_cnt", "call_repressed_cnt"),
                       labels=c("Up-regulation calls", "Down-regulation calls")) +
  theme(legend.position="bottom") +
  ggsave("./imgs/repeated-sig-regulate-counts.png", width=6, height=4)
```

This next graph is based on count of down-regulation and up-regulation edges in evolved gene regulatory networks (where vertices are modules) that solve the repeated signal task.

```{r}
alt_sig_network_data_loc <- paste(data_dir, "alt_sig_reg_graphs_summary.csv", sep="")
alt_sig_network_data <- read.csv(alt_sig_network_data_loc, na.strings="NA")

temp <- filter(max_fit_org_data, 
               solution=="1" & condition=="both")
melted <- melt(filter(alt_sig_network_data, 
                      run_id %in% temp$SEED), 
               variable.name = "reg_edge_type",
               value.name = "reg_edges_cnt",
               measure.vars=c("repressed_edges_cnt", "promoted_edges_cnt"))
get_num_sig_resps <- function(seed) {
  return(filter(temp, SEED==seed)$NUM_SIGNAL_RESPONSES)
}
melted$NUM_SIGNAL_RESPONSES <- mapply(get_num_sig_resps, 
                                      melted$run_id)
melted$NUM_SIGNAL_RESPONSES <- factor(melted$NUM_SIGNAL_RESPONSES)

ggplot(melted, aes(x=NUM_SIGNAL_RESPONSES, y=reg_edges_cnt, color=reg_edge_type)) +
  geom_boxplot() +
  xlab("Environmental Complexity") +
  ylab("# Edges") +
  scale_color_discrete(name="Edge type:", 
                       limits=c("repressed_edges_cnt", "promoted_edges_cnt"),
                       labels=c("Repressing edges", "Promoting edges")) +
  theme(legend.position="bottom") +
  ggsave("./imgs/repeated-signal-regulation-edges.png", width=5, height=4)
```

## Case Study: regulation in action

```{r}
trace_id <- 100234
```

Next, we take a closer look at the behavioral profile of evolved SignalGP organisms.

We'll be taking a closer look at solution from run #`r trace_id`, which evolved in a 4-state environment that with access regulation.

```{r}
# Extract relevant information about solution of interest.
org_info <- filter(max_fit_org_data, 
                   SEED==trace_id)
num_envs <- org_info$NUM_SIGNAL_RESPONSES
score <- org_info$score
condition <- org_info$condition
is_sol <- org_info$solution
num_modules <- org_info$num_modules
# Load trace file associated with this solution.
trace_file <- paste(data_dir, "traces/trace_update-10000_run-id-", 
                    trace_id, ".csv", sep="")
trace_data <- read.csv(trace_file, na.strings="NONE")
trace_data$similarity_score <- 1 - trace_data$match_score
trace_data$is_running <- trace_data$is_running > 0
trace_data$regulator_state_simplified <- NA

# infer which modules were triggered
trace_data$triggered <- (trace_data$env_signal_closest_match == trace_data$module_id) &
                          (trace_data$cpu_step == "0")

categorize_reg_state <- function(reg_state) {
  if (reg_state == 0) {
    return("neutral")
  } else if (reg_state < 0) {
    return("promoted")
  } else if (reg_state > 0) {
    return("repressed")
  } else {
    return("unknown")
  }
}
trace_data$regulator_state_simplified <- mapply(categorize_reg_state,
                                                trace_data$regulator_state)

# Extract only rows that correspond with modules that were active during evaluation.
active_data <- filter(trace_data, is_ever_active=="1")
# Do some work to have module ids appear in a nice order along axis.
active_module_ids <- levels(factor(active_data$module_id))
active_module_ids <- as.integer(active_module_ids)
module_id_map <- as.data.frame(active_module_ids)
module_id_map$order <- order(module_id_map$active_module_ids) - 1
get_module_x_pos <- function(module_id) {
  return(filter(module_id_map, active_module_ids==module_id)$order)
}
active_data$mod_id_x_pos <- mapply(get_module_x_pos, active_data$module_id)

trace_data <- active_data # omit all non-active modules from visualization
```

### Regulator state over time.

```{r}
out_name <- paste("./imgs/case-study-trace-id-", trace_id, "-regulator-state-ao.pdf", sep="")
reg_plot <- ggplot(trace_data, 
                   aes(x=mod_id_x_pos, 
                       y=time_step, 
                       fill=regulator_state_simplified)) +
  scale_fill_viridis(name="State:",
                     limits=c("promoted", "neutral", "repressed"),
                     labels=c("Promoted", "Neutral", "Repressed"),
                     discrete=TRUE, 
                     direction=-1) +
  scale_x_discrete(name="Function ID",
                 limits=seq(0, length(active_module_ids)-1, 1),
                 labels=active_module_ids) +
  scale_y_discrete(name="Time Step",
                   limits=seq(0, 50, 10)) +
  # Background
  geom_tile(color="white", 
            size=0.2, 
            width=1, 
            height=1) +
  # Module is-running highlights
  geom_tile(data=filter(trace_data, is_running==TRUE | triggered==TRUE), 
            color="black", 
            width=1, 
            height=1, 
            size=0.8) +
  # Environment delimiters
  geom_hline(yintercept=filter(trace_data, cpu_step==0)$time_step - 0.5, 
             size=1) +
  # Draw points on triggered modules
  geom_point(data=filter(trace_data, triggered==TRUE),
             shape=21, colour="black", fill="white", stroke=0.33, size=0.75,
             position=position_nudge(x = 0, y = 0.01)) +
  theme(legend.position = "top",
        legend.text = element_text(size=10),
        axis.text.y = element_text(size=10),
        axis.text.x = element_text(size=10)) +
  ggtitle("Function Regulation States") +
  coord_flip() +
  ggsave(out_name, height=3, width=4)
reg_plot
```

### Function-signal tag-match score over time

```{r}

out_name <- paste("./imgs/case-study-trace-id-", trace_id, "-similarity-score-ao.pdf", sep="")
match_plot <- ggplot(trace_data, 
                     aes(x=mod_id_x_pos, 
                         y=time_step, 
                         fill=similarity_score)) +
  scale_fill_viridis(option="plasma",
                     name="Score:  ",
                     limits=c(0, 1.0),
                     breaks=c(0, 0.25, 0.50, 0.75, 1.0),
                     labels=c("0%", "25%", "50%", "75%", "100%")) +
  scale_x_discrete(name="Function ID",
                   limits=seq(0, length(active_module_ids)-1, 1),
                   labels=active_module_ids) +
  scale_y_discrete(name="Time Step",
                   limits=seq(0, 50, 10)) +
  # Background
  geom_tile(color="white", 
            size=0.2, 
            width=1, 
            height=1) +
  # Module is-running highlights
  geom_tile(data=filter(trace_data, is_running==TRUE | triggered==TRUE), 
            color="black", 
            width=1, 
            height=1, 
            size=0.8) +
  # Environment delimiters
  geom_hline(yintercept=filter(trace_data, cpu_step==0)$time_step-0.5, size=1) +
  # Draw points on triggered modules
  geom_point(data=filter(trace_data, triggered==TRUE),
             shape=21, colour="black", fill="white", stroke=0.33, size=0.75,
             position=position_nudge(x = 0, y = 0.01)) +
  theme(legend.position = "top",
        legend.text = element_text(size=10),
        axis.text.y = element_text(size=10),
        axis.text.x = element_text(size=10)) +
  guides(fill = guide_colourbar(barwidth = 10, barheight = 0.5)) +
  ggtitle("Function Match Scores") +
  coord_flip() +
  ggsave(out_name, height=3, width=4)
match_plot
```

### Evolved regulatory network

```{r}
# Networks!
graph_nodes_loc <- paste(data_dir, 
                         "reg-igraphs/reg_graph_id-", trace_id, "_nodes.csv", sep="")
graph_edges_loc <- paste(data_dir,
                         "reg-igraphs/reg_graph_id-", trace_id, "_edges.csv", sep="")
graph_nodes_data <- read.csv(graph_nodes_loc, na.strings="NONE")
graph_edges_data <- read.csv(graph_edges_loc, na.strings="NONE")

network <- graph_from_data_frame(d=graph_edges_data, 
                                 vertices=graph_nodes_data, 
                                 directed=TRUE)

# Setup edge styling
E(network)$color[E(network)$type == "promote"] <- "#FCE640"
E(network)$lty[E(network)$type == "promote"] <- 1
E(network)$color[E(network)$type == "repress"] <- "#441152"
E(network)$lty[E(network)$type == "repress"] <- 2

output_file <- FALSE
network_out_name <- paste("./imgs/case-study-id-", 
                          trace_id, 
                          "-network.svg", sep="")
if (output_file) {
  svg(network_out_name, width=4.5,height=4)
  # bottom, left, top, right
  # par(mar=c(0.2,0,1,0))
}
plot(network, 
     main="Regulation Network",
     edge.arrow.size=.5, 
     edge.width=2,
     vertex.size=20,
     curved=TRUE,
     vertex.color="grey99",
     vertex.label.color="black",
     vertex.label.family="sans",
     layout=layout.auto(network))
legend(x = "bottomleft",      ## position, also takes x,y coordinates
       legend = c("Promoted", "Repressed"),
       pch = 19,              ## legend symbols see ?points
       col = c("#FCE640", "#441152"),
       bty = "n",
       title = "Edges")
if (output_file) {
  dev.flush()
  dev.off()
}
```