# Boolean calculator problem (postfix notation)

Here, we give an overview of the boolean-logic calculator problem, and we provide our data analyses for related experiments.
All of our source code for statistical analyses and data visualizations is embedded in this document.
The raw data can be found on the OSF project associated with this work [@Lalejini_Moreno_Ofria_Data_2020].

**Please [file an issue or make a pull request on github](https://github.com/amlalejini/Tag-based-Genetic-Regulation-for-LinearGP/issues) to report any mistakes, ask questions, request more explanation, et cetera.**

## Overview

```{r}
# Experimental parameters referenced in-text all in one convenient place.
time_steps <- 128
replicates <- 200
population_size <- 1000
generations <- 10000

# Settings for statistical analyses.
alpha <- 0.05

# Relative location of data.
working_directory <- "experiments/2020-11-28-bool-calc-postfix/analysis/" # << For bookdown
# working_directory <- "./"                                              # << For local analysis
```

We use a modified version of the Boolean-logic calculator problem to further investigate the potential for our implementation of tag-based regulation to impede adaptive evolution.
Our previous experiments with the Boolean-logic calculator problem provided inputs in prefix notation: the operator (e.g., AND, OR, XOR, etc.) is specified first, followed by the requisite number of numeric operands.
As such, the final input signal does not differentiate which type of computation a program is expected to perform (e.g., AND, OR, XOR, etc.).
This requires programs to adjust their response to the final input signal based on the context provided by the previous two signals, thereby increasing the value of regulation.

We explore whether the calculator problem's context-dependence is driving the benefit of tag-based regulation that we identified in previous experiments.
We can reduce context-dependence of the calculator problem by presenting input sequences in postfix notation.
In postfix notation, programs receive the requisite numeric operand inputs first and the operator input last.
As such, the final signal in an input sequence will always differentiate which bitwise operation should be performed.
Successful programs must store the numeric inputs embedded in operand signals, and then, as in the changing-signal problem, a distinct signal will differentiate which of the response types a program should execute.


## Analysis Dependencies

Load all required R libraries.

```{r, message=FALSE}
library(ggplot2)
library(tidyverse)
library(cowplot)
library(viridis)
library(reshape2)
library(igraph)
source("https://gist.githubusercontent.com/benmarwick/2a1bb0133ff568cbe28d/raw/fb53bd97121f7f9ce947837ef1a4c65a73bffb3f/geom_flat_violin.R")
```

These analyses were conducted in the following computing environment:

```{r}
print(version)
```

## Setup

Load data, initial data cleanup, configure some global settings.

```{r}
data_loc <- paste0(working_directory, "data/max_fit_orgs.csv")
data <- read.csv(data_loc, na.strings="NONE")

# Specify factors (not all of these matter for this set of runs).
data$matchbin_thresh <- factor(
  data$matchbin_thresh,
  levels=c(0, 25, 50, 75)
)

data$TAG_LEN <- factor(
  data$TAG_LEN,
  levels=c(32, 64, 128, 256)
)

data$notation <- factor(
  data$notation,
  levels=c("prefix", "postfix")
)

# Define function to summarize regulation/memory configurations.
get_con <- function(reg, mem) {
  if (reg == "0" && mem == "0") {
    return("none")
  } else if (reg == "0" && mem=="1") {
    return("memory")
  } else if (reg=="1" && mem=="0") {
    return("regulation")
  } else if (reg=="1" && mem=="1") {
    return("both")
  } else {
    return("UNKNOWN")
  }
}

# Specify experimental condition for each datum.
data$condition <- mapply(
  get_con,
  data$USE_FUNC_REGULATION,
  data$USE_GLOBAL_MEMORY
)

data$condition <- factor(
  data$condition,
  levels=c("regulation", "memory", "none", "both")
)

# Given knockout info, what strategy does a program use?
get_strategy <- function(use_reg, use_mem) {
  if (use_reg=="0" && use_mem=="0") {
    return("use neither")
  } else if (use_reg=="0" && use_mem=="1") {
    return("use memory")
  } else if (use_reg=="1" && use_mem=="0") {
    return("use regulation")
  } else if (use_reg=="1" && use_mem=="1") {
    return("use both")
  } else {
    return("UNKNOWN")
  }
}

# Specify experimental conditions (to make labeling easier).
data$strategy <- mapply(
  get_strategy,
  data$relies_on_regulation,
  data$relies_on_global_memory
)

data$strategy <- factor(
  data$strategy,
  levels=c(
    "use regulation",
    "use memory",
    "use neither",
    "use both"
  )
)

# Filter data to include only replicates labeled as solutions
sol_data <- filter(data, solution=="1")

####### Load instruction execution data #######
inst_exec_data <- read.csv(paste0(working_directory, "data/exec_trace_summary.csv"), na.strings="NA")

inst_exec_data$condition <- mapply(
  get_con,
  inst_exec_data$USE_FUNC_REGULATION,
  inst_exec_data$USE_GLOBAL_MEMORY
)

inst_exec_data$condition <- factor(
  inst_exec_data$condition,
  levels=c("regulation", "memory", "none", "both")
)

inst_exec_data$notation <- factor(
  inst_exec_data$notation,
  levels=c("prefix", "postfix")
)

####### Load network data #######
reg_network_data <- read.csv(paste0(working_directory, "data/reg_graphs_summary.csv"), na.strings="NA")
reg_network_data <- filter(reg_network_data, run_id %in% data$SEED)

get_notation <- function(seed) {
  return(filter(data, SEED==seed)$notation)
}

reg_network_data$notation <- mapply(
  get_notation,
  reg_network_data$run_id
)

reg_network_data$notation <- factor(
  reg_network_data$notation,
  levels=c("prefix", "postfix")
)

####### misc #######
# Configure our default graphing theme
theme_set(theme_cowplot())
```

## Problem-solving success

The number of successful replicates by condition:

```{r, echo=TRUE}
# Graph the number of solutions evolved in each condition, faceted by environmental complexity
ggplot(sol_data, aes(x=condition, fill=condition)) +
  geom_bar() +
  geom_text(
    stat="count",
    mapping=aes(label=..count..),
    position=position_dodge(0.9),
    vjust=0
  ) +
  scale_x_discrete(
    name="Condition",
    breaks=c("memory","both"),
    labels=c("Regulation-\ndisabled", "Regulation-\nenabled")
  ) +
  ylab("Successful replciates") +
  ylim(0, 200) +
  theme(legend.position = "none") +
  ggsave(
    paste0(working_directory, "imgs/boolean-calc-postfix-solution-counts.pdf"),
    width=4,
    height=4
  )
```

Test for significance using Fisher's exact test.

```{r}
# Extract successes/fails for each condition.
reg_disabled_success_cnt <- nrow(filter(sol_data, solution=="1" & condition=="memory"))
reg_disabled_fail_cnt <- replicates - reg_disabled_success_cnt

reg_enabled_success_cnt <- nrow(filter(sol_data, solution=="1" & condition=="both"))
reg_enabled_fail_cnt <- replicates - reg_enabled_success_cnt

# Regulation-disabled vs regulation-enabled
perf_table <- matrix(
  c(
    reg_enabled_success_cnt,
    reg_disabled_success_cnt,
    reg_enabled_fail_cnt,
    reg_disabled_fail_cnt
    ),
    nrow=2
)

rownames(perf_table) <- c("reg-enabled", "reg-disabled")
colnames(perf_table) <- c("success", "fail")

print(perf_table)
print(fisher.test(perf_table))
```

## How many generations elapse before solutions evolve?

```{r}
unfinished_data <- filter(data, solution=="0")
unfinished_data$graph_update <- 12500

ggplot( ) +
  geom_flat_violin(
    data = sol_data,
    mapping = aes(x=condition, y=update, fill=condition),
    position = position_nudge(x = .2, y = 0),
    alpha = .8
  ) +
  geom_point(
    data = sol_data,
    aes(x=condition, y=update, color=condition),
    position = position_jitter(width = .15),
    size = .5,
    alpha = 0.8
  ) +
  geom_point(
    data = unfinished_data,
    mapping=aes(x=condition, y=graph_update),
    color="gray",
    position = position_jitter(width = .15, height=500),
    size = .5,
    alpha = 0.8
  ) +
  geom_boxplot(
    data = sol_data,
    mapping = aes(x=condition, y=update, fill=condition),
    width = .1,
    outlier.shape = NA,
    alpha = 0.5
  ) +
  scale_x_discrete(
    name="Condition",
    breaks=c("memory", "both"),
    labels=c("Regulation-\ndisabled", "Regulation-\nenabled")
  ) +
  scale_y_continuous(
    name="Generation first solution evolved",
    limits=c(0, 13000),
    breaks=c(0, 2500, 5000, 7500, 10000, 12500),
    labels=c("0", "2500", "5000", "7500", "10000", "Unsolved")
  ) +
  # coord_flip() +
  guides(fill = FALSE) +
  guides(color = FALSE) +
  ggsave(
    paste0(working_directory, "imgs/boolean-calc-postfix-solve-time-cloud.pdf"),
    width=4,
    height=4
  )
```

Test for statistical difference between conditions using a Wilcoxon rank sum test.

```{r}
print(wilcox.test(formula=update~condition, data=sol_data, exact=FALSE, conf.int=TRUE))
```

## Evolved strategies

### What mechanisms do programs rely on to adjust responses to signals over time?

We used indpendent knockouts of tag-based genetic regulation and global memory buffer access to investigate the mechanisms underpinning successful programs.

```{r}
ggplot( sol_data, mapping=aes(x=condition, fill=strategy) ) +
  geom_bar(
    position="fill",
    stat="count"
  ) +
  geom_text(
    stat='count',
    mapping=aes(label=..count..),
    position=position_fill(vjust=0.05)
  ) +
  ylab("% of Solutions") +
  scale_fill_discrete(
    name="Strategy:",
    breaks=c(
      "use regulation",
      "use memory",
      "use neither",
      "use both"
    ),
    labels=c(
      "Use regulation (only)",
      "Use global memory (only)",
      "Use neither",
      "Use both"
    )
  ) +
  scale_x_discrete(
    name="Condition",
    breaks=c("memory", "both"),
    labels=c("Regulation-\ndisabled", "Regulation-\nenabled")
  ) +
  theme(legend.position = "bottom")
```

### Gene regulatory networks

Looking only at successful programs that rely on regulation. At a glance, what do gene regulatory networks look like?

First, the total edges found in networks:

```{r}
relies_on_reg <- filter(
  sol_data,
  relies_on_regulation=="1"
)$SEED

ggplot( filter(reg_network_data, run_id %in% relies_on_reg), aes(x=notation, y=edge_cnt) ) +
  geom_boxplot() +
  geom_jitter(alpha=0.1) +
  xlab("Notation") +
  ylab("# Edges") +
  theme(
    legend.position="bottom",
    legend.text=element_text(size=9),
    legend.title=element_text(size=10),
    axis.title.x=element_text(size=12)
  ) +
  ggsave(
    paste0(working_directory, "imgs/boolean-calc-postfix-regulation-edges.png"),
    width=4,
    height=3
  )
```


Next, let's look at edges by type.

```{r}

# Process/cleanup the network data
melted_network_data <- melt(
  filter(reg_network_data,
         run_id %in% relies_on_reg
        ),
  variable.name = "reg_edge_type",
  value.name = "reg_edges_cnt",
  measure.vars=c("repressed_edges_cnt", "promoted_edges_cnt")
)

ggplot( melted_network_data, aes(x=reg_edge_type, y=reg_edges_cnt, color=reg_edge_type) ) +
  geom_boxplot() +
  geom_jitter(alpha=0.2) +
  xlab("Environmental Complexity") +
  ylab("# Edges") +
  scale_x_discrete(
    name="Edge type",
    limits=c("repressed_edges_cnt", "promoted_edges_cnt"),
    labels=c("Repressing edges", "Promoting edges")
  ) +
  theme(
    legend.position="none",
    legend.text=element_text(size=9),
    legend.title=element_text(size=10),
    axis.title.x=element_text(size=12)
  ) +
  ggsave(
    paste0(working_directory, "imgs/boolean-calc-postfix-regulation-edge-types.png"),
    width=4,
    height=3
  )
```

Test for a statistical difference between edge types using a wilcoxon rank sum test:

```{r}
print(
  paste0(
    "Median # repressed edges: ",
    median(filter(melted_network_data, reg_edge_type=="repressed_edges_cnt")$reg_edges_cnt)
  )
)

print(
  paste0(
    "Median # promoting edges: ",
    median(filter(melted_network_data, reg_edge_type=="promoted_edges_cnt")$reg_edges_cnt)
  )
)

print(wilcox.test(formula=reg_edges_cnt ~ reg_edge_type, data=melted_network_data, exact=FALSE, conf.int=TRUE))
```

### Program instruction execution traces

#### Execution time

How many time steps do successful programs take to solve the boolean calculator problem?

```{r}
# only want solutions
solutions_inst_exec_data <- filter(inst_exec_data, SEED %in% sol_data$SEED)

ggplot( solutions_inst_exec_data, aes(x=condition, y=total_execution_time, color=condition) ) +
  geom_boxplot() +
  geom_jitter(alpha=0.2) +
  scale_x_discrete(
    breaks=c("memory", "both"),
    labels=c("Regulation-\ndisabled", "Regulation-\nenabled")
  ) +
  theme(
    legend.position="none"
  )
```

Test for significant difference between conditions using Wilcoxon rank sum test:

```{r}
print(
  wilcox.test(
    formula=total_execution_time~condition,
    data=filter(solutions_inst_exec_data),
    exact=FALSE,
    conf.int=TRUE)
)
```

#### What types of instructions to successful programs execute?

Here, we look at the distribution of instruction types executed by successful programs.
We're primarily interested in the proportion of control flow instructions, so let's look at that first.

```{r}
ggplot( solutions_inst_exec_data, aes(x=condition, y=control_flow_inst_prop, color=condition) ) +
  geom_boxplot() +
  geom_jitter(alpha=0.2) +
  scale_x_discrete(
    breaks=c("memory", "both"),
    labels=c("Regulation-\ndisabled", "Regulation-\nenabled")
  ) +
  ylab("Proportion of executed flow control instructions") +
  theme(
    legend.position="bottom",
    axis.title.x=element_blank()
  )
```

Test for significant difference between conditions using a Wilcoxon rank sum test:

```{r}
print(
  wilcox.test(
    formula=control_flow_inst_prop~condition,
    data=filter(solutions_inst_exec_data),
    exact=FALSE,
    conf.int=TRUE)
)
```

In case you're curious, here's all categories of instructions:

```{r}
melted <- melt(
  solutions_inst_exec_data,
  variable.name = "inst_type",
  value.name = "inst_type_prop",
  measure.vars=c(
    "math_inst_prop",
    "module_inst_prop",
    "memory_inst_prop",
    "regulation_inst_prop",
    "control_flow_inst_prop",
    "thread_inst_prop",
    "task_inst_prop",
    "nop_inst_prop"
  )
)

ggplot( melted, aes(x=inst_type, y=inst_type_prop, color=condition) ) +
  geom_boxplot() +
  scale_color_discrete(
    name="Condition:",
    breaks=c("memory", "both"),
    labels=c("Regulation-disabled", "Regulation-enabled")
  ) +
  xlab("Instruction type") +
  ylab("Proportion of instructions in execution trace") +
  coord_flip() +
  theme(legend.position="bottom")
```

## Visualizaing an evolved regulatory network

Let's take a closer look at a successful gene regulatory network.

```{r}
trace_id <- 25392
```

Specifically, we'll be looking at the solution evolved in run id `r trace_id` (arbitrarily selected).


### Evolved regulatory network

We use the igraph package to draw this program's gene regulatory network.

```{r, message=FALSE}
# Networks!
graph_nodes_loc <- paste0(working_directory, "data/igraphs/reg_graph_id-", trace_id, "_nodes.csv")
graph_edges_loc <- paste0(working_directory, "data/igraphs/reg_graph_id-", trace_id, "_edges.csv")
graph_nodes_data <- read.csv(graph_nodes_loc, na.strings="NONE")
graph_edges_data <- read.csv(graph_edges_loc, na.strings="NONE")

network <- graph_from_data_frame(
  d=graph_edges_data,
  vertices=graph_nodes_data,
  directed=TRUE
)

# Setup edge styling
E(network)$color[E(network)$type == "promote"] <- "#FCE640"
E(network)$lty[E(network)$type == "promote"] <- 1
E(network)$color[E(network)$type == "repress"] <- "#441152"
E(network)$lty[E(network)$type == "repress"] <- 1

network_out_name <- paste0(working_directory, "imgs/case-study-id-", trace_id, "-network.svg")

draw_network <- function(net, write_out, out_name) {
  if (write_out) {
    svg(out_name, width=4,height=1.5)
    # bottom, left, top, right
    par(mar=c(0.2,0,1,0.5))
  }
  plot(
    net,
    edge.arrow.size=0.4,
    edge.arrow.width=0.75,
    edge.width=2,
    vertex.size=10,
    vertex.label.cex=0.65,
    curved=TRUE,
    vertex.color="grey99",
    vertex.label.color="black",
    vertex.label.family="sans",
    layout=layout.circle(net)
  )
  legend(
    x = "bottomleft",      ## position, also takes x,y coordinates
    legend = c("Promoted", "Repressed"),
    pch = 19,              ## legend symbols see ?points
    col = c("#FCE640", "#441152"),
    bty = "n",
    border="black",
    xpd=TRUE,
    title = "Edges"
  )
  if (write_out) {
    dev.flush()
    dev.off()
  }
}

draw_network(network, TRUE, network_out_name)
draw_network(network, FALSE, "")
```

