# Contextual-signal problem analysis

Here, we give an overview of the contextual-signal diagnostic problem, and we provide our data analyses for related experiments.
All of our source code for statistical analyses and data visualizations is embedded in this document.
The raw data can be found on the OSF project associated with this work (link coming).

**Please [file an issue or make a pull request on github](https://github.com/amlalejini/Tag-based-Genetic-Regulation-for-LinearGP/issues) to report any mistakes, ask questions, request more explanation, et cetera.**

## Overview

```{r}
# Experimental parameters referenced in-text all in one convenient place.
time_steps <- 128
replicates <- 200
population_size <- 1000
generations <- 10000

# Settings for statistical analyses.
alpha <- 0.05

# Relative location of data.
working_directory <- "experiments/2020-11-27-context-sig/analysis/" # << For bookdown
# working_directory <- "./"                                         # << For local analysis
```

In the contextual-signal problem, programs must respond appropriately to a sequence of two input signals where the first, ''contextual'', signal dictates how a program should respond to each possible second, ''response'', signal.
In this work, there are a total of four possible input signals and four possible output responses.
Programs output these responses by executing one of four response instructions.

The dataframe below gives the correct output for each combination of input signals.
```{r}
testcases <- read.csv(paste0(working_directory, "../hpcc/examples_S4.csv"))
print(testcases)
```

## Analysis Dependencies

Load all required R libraries.

```{r, message=FALSE}
library(ggplot2)
library(tidyverse)
library(cowplot)
library(viridis)
library(reshape2)
library(igraph)
source("https://gist.githubusercontent.com/benmarwick/2a1bb0133ff568cbe28d/raw/fb53bd97121f7f9ce947837ef1a4c65a73bffb3f/geom_flat_violin.R")
```

These analyses were conducted in the following computing environment:

```{r}
print(version)
```

## Setup

Load data, initial data cleanup, configure some global settings.

```{r}
####### Load max fit program data #######
data_loc <- paste0(working_directory, "data/max_fit_orgs.csv")
data <- read.csv(data_loc, na.strings="NONE")

# Specify factors (not all of these matter for this set of runs).
data$matchbin_thresh <- factor(
  data$matchbin_thresh,
  levels=c(0, 25, 50, 75)
)

data$TAG_LEN <- factor(
  data$TAG_LEN,
  levels=c(32, 64, 128, 256)
)

data$task <- factor(
  data$task,
  levels=c("S2", "S3", "S4")
)

# Filter down to only data we use in paper.
data <- filter(data, task=="S4")

# Define function to summarize regulation/memory configurations.
get_con <- function(reg, mem) {
  if (reg == "0" && mem == "0") {
    return("none")
  } else if (reg == "0" && mem=="1") {
    return("memory")
  } else if (reg=="1" && mem=="0") {
    return("regulation")
  } else if (reg=="1" && mem=="1") {
    return("both")
  } else {
    return("UNKNOWN")
  }
}
# Specify experimental condition for each datum.
data$condition <- mapply(
  get_con,
  data$USE_FUNC_REGULATION,
  data$USE_GLOBAL_MEMORY
)

data$condition <- factor(
  data$condition,
  levels=c("regulation", "memory", "none", "both")
)

# Given knockout info, what strategy does a program use?
get_strategy <- function(use_reg, use_mem) {
  if (use_reg=="0" && use_mem=="0") {
    return("use neither")
  } else if (use_reg=="0" && use_mem=="1") {
    return("use memory")
  } else if (use_reg=="1" && use_mem=="0") {
    return("use regulation")
  } else if (use_reg=="1" && use_mem=="1") {
    return("use both")
  } else {
    return("UNKNOWN")
  }
}

# Specify experimental conditions (to make labeling easier).
data$strategy <- mapply(
  get_strategy,
  data$relies_on_regulation,
  data$relies_on_global_memory
)

data$strategy <- factor(
  data$strategy,
  levels=c(
    "use regulation",
    "use memory",
    "use neither",
    "use both"
  )
)

# Filter data to include only replicates labeled as solutions
sol_data <- filter(data, solution=="1")

####### Load instruction execution data #######
inst_exec_data <- read.csv(paste0(working_directory, "data/exec_trace_summary.csv"), na.strings="NA")

inst_exec_data$condition <- mapply(
  get_con,
  inst_exec_data$USE_FUNC_REGULATION,
  inst_exec_data$USE_GLOBAL_MEMORY
)

inst_exec_data$condition <- factor(
  inst_exec_data$condition,
  levels=c("regulation", "memory", "none", "both")
)

inst_exec_data$task <- factor(
  inst_exec_data$task,
  levels=c("S2", "S3", "S4")
)

####### Load network data #######
reg_network_data <- read.csv(paste0(working_directory, "data/reg_graphs_summary.csv"), na.strings="NA")
reg_network_data <- filter(reg_network_data, run_id %in% data$SEED)

get_task <- function(seed) {
  return(filter(data, SEED==seed)$task)
}

reg_network_data$task <- mapply(
  get_task,
  reg_network_data$run_id
)

reg_network_data$task <- factor(reg_network_data$task)

####### misc #######
# Configure our default graphing theme
theme_set(theme_cowplot())
```

## Problem-solving success

The number of successful replicates by condition:

```{r, echo=TRUE}
# Graph the number of solutions evolved in each condition, faceted by environmental complexity
ggplot(filter(sol_data, task=="S4"), aes(x=condition, fill=condition)) +
  geom_bar() +
  geom_text(
    stat="count",
    mapping=aes(label=..count..),
    position=position_dodge(0.9),
    vjust=0
  ) +
  scale_x_discrete(
    name="Condition",
    breaks=c("memory","both"),
    labels=c("Regulation-\ndisabled", "Regulation-\nenabled")
  ) +
  ylab("# successful replciates") +
  theme(legend.position = "none") +
  ggsave(
    paste0(working_directory, "imgs/context-signal-solution-counts.pdf"),
    width=4,
    height=4
  )
```

Test for significance using Fisher's exact test.

```{r}
# Extract successes/fails for each condition.
reg_disabled_success_cnt <- nrow(filter(sol_data, task=="S4" & solution=="1" & condition=="memory"))
reg_disabled_fail_cnt <- replicates - reg_disabled_success_cnt

reg_enabled_success_cnt <- nrow(filter(sol_data, task=="S4" & solution=="1" & condition=="both"))
reg_enabled_fail_cnt <- replicates - reg_enabled_success_cnt

# Regulation-disabled vs regulation-enabled
perf_table <- matrix(
  c(
    reg_enabled_success_cnt,
    reg_disabled_success_cnt,
    reg_enabled_fail_cnt,
    reg_disabled_fail_cnt
    ),
    nrow=2
)

rownames(perf_table) <- c("reg-enabled", "reg-disabled")
colnames(perf_table) <- c("success", "fail")

print(perf_table)
print(fisher.test(perf_table))
```

## How many generations elapse before solutions evolve?

```{r}
ggplot( data = filter(sol_data, task=="S4"), aes(x=condition, y=update, fill=condition) ) +
  geom_flat_violin(
    position=position_nudge(x = .2, y = 0),
    alpha=.8
  ) +
  geom_point(
    aes(y=update, color=condition),
    position = position_jitter(width = .15),
    size = .5,
    alpha = 0.8
  ) +
  geom_boxplot(
    width = .1,
    outlier.shape = NA,
    alpha = 0.5
  ) +
  scale_x_discrete(
    name="Condition",
    breaks=c("memory", "both"),
    labels=c("Regulation-\ndisabled", "Regulation-\nenabled")
  ) +
  scale_y_continuous(name="Generation first solution evolved") +
  guides(fill = FALSE) +
  guides(color = FALSE) +
  ggsave(
    paste0(working_directory, "imgs/context-signal-solve-time-cloud.png"),
    width=4,
    height=4
  )
```

Test for statistical difference between conditions using a Wilcoxon rank sum test.

```{r}
print(wilcox.test(formula=update~condition, data=filter(sol_data, task=="S4"), exact=FALSE, conf.int=TRUE))
```

## Evolved strategies

### What mechanisms do programs rely on to adjust responses to signals over time?

We used indpendent knockouts of tag-based genetic regulation and global memory buffer access to investigate the mechanisms underpinning successful programs.

```{r}
ggplot( filter(sol_data, task=="S4"), mapping=aes(x=condition, fill=strategy) ) +
  geom_bar(
    position="fill",
    stat="count"
  ) +
  geom_text(
    stat='count',
    mapping=aes(label=..count..),
    position=position_fill(vjust=0.05)
  ) +
  ylab("% of Solutions") +
  scale_fill_discrete(
    name="Strategy:",
    breaks=c(
      "use regulation",
      "use memory",
      "use neither",
      "use both"
    ),
    labels=c(
      "Use regulation (only)",
      "Use global memory (only)",
      "Use neither",
      "Use both"
    )
  ) +
  scale_x_discrete(
    name="Condition",
    breaks=c("memory", "both"),
    labels=c("Regulation-\ndisabled", "Regulation-\nenabled")
  ) +
  theme(legend.position = "bottom")
```

### Gene regulatory networks

Looking only at successful programs that rely on regulation. At a glance, what do gene regulatory networks look like?

First, the total edges found in networks:

```{r}
relies_on_reg <- filter(
  sol_data,
  relies_on_regulation=="1"
)$SEED

ggplot( filter(reg_network_data, run_id %in% relies_on_reg & task=="S4"), aes(x=task, y=edge_cnt) ) +
  geom_boxplot() +
  geom_jitter(alpha=0.1) +
  xlab("Task") +
  ylab("# Edges") +
  theme(
    legend.position="bottom",
    legend.text=element_text(size=9),
    legend.title=element_text(size=10),
    axis.title.x=element_text(size=12)
  ) +
  ggsave(
    paste0(working_directory, "imgs/contextual-signal-regulation-edges.png"),
    width=4,
    height=3
  )
```

Next, let's look at edges by type.

```{r}

# Process/cleanup the network data
melted_network_data <- melt(
  filter(reg_network_data,
         run_id %in% relies_on_reg
        ),
  variable.name = "reg_edge_type",
  value.name = "reg_edges_cnt",
  measure.vars=c("repressed_edges_cnt", "promoted_edges_cnt")
)

ggplot(filter(melted_network_data, task=="S4"), aes(x=reg_edge_type, y=reg_edges_cnt, color=reg_edge_type)) +
  geom_boxplot() +
  geom_jitter(alpha=0.2) +
  xlab("Environmental Complexity") +
  ylab("# Edges") +
  scale_x_discrete(
    name="Edge type",
    limits=c("repressed_edges_cnt", "promoted_edges_cnt"),
    labels=c("Repressing edges", "Promoting edges")
  ) +
  theme(
    legend.position="none",
    legend.text=element_text(size=9),
    legend.title=element_text(size=10),
    axis.title.x=element_text(size=12)
  ) +
  ggsave(
    paste0(working_directory, "imgs/context-signal-regulation-edge-types.png"),
    width=4,
    height=3
  )
```

Test for a statistical difference between edge types using a wilcoxon rank sum test:

```{r}
print(
  paste0(
    "Median # repressed edges: ",
    median(filter(melted_network_data, task=="S4" & reg_edge_type=="repressed_edges_cnt")$reg_edges_cnt)
  )
)

print(
  paste0(
    "Median # promoting edges: ",
    median(filter(melted_network_data, task=="S4" & reg_edge_type=="promoted_edges_cnt")$reg_edges_cnt)
  )
)

print(wilcox.test(formula=reg_edges_cnt ~ reg_edge_type, data=filter(melted_network_data, task=="S4"), exact=FALSE, conf.int=TRUE))

```

### Program instruction execution traces

#### Execution time

How many time steps do evolved programs use to solve the contextual-signal task?

```{r}
# only want solutions
solutions_inst_exec_data <- filter(inst_exec_data, SEED %in% sol_data$SEED & task=="S4")

ggplot( solutions_inst_exec_data, aes(x=condition, y=total_execution_time, color=condition) ) +
  geom_boxplot() +
  geom_jitter(alpha=0.2) +
  scale_x_discrete(
    breaks=c("memory", "both"),
    labels=c("Regulation-\ndisabled", "Regulation-\nenabled")
  ) +
  theme(
    legend.position="none"
  )
```

Test for significant difference between conditions using Wilcoxon rank sum test:

```{r}
print(
  wilcox.test(
    formula=total_execution_time~condition,
    data=filter(solutions_inst_exec_data),
    exact=FALSE,
    conf.int=TRUE)
)
```

#### What types of instructions to successful programs execute?

Here, we look at the distribution of instruction types executed by successful programs.
We're primarily interested in the proportion of control flow instructions, so let's look at that first.

```{r}
ggplot( solutions_inst_exec_data, aes(x=condition, y=control_flow_inst_prop, color=condition) ) +
  geom_boxplot() +
  geom_jitter(alpha=0.2) +
  scale_x_discrete(
    breaks=c("memory", "both"),
    labels=c("Regulation-\ndisabled", "Regulation-\nenabled")
  ) +
  ylab("Proportion of executed flow control instructions") +
  theme(
    legend.position="bottom",
    axis.title.x=element_blank()
  )
```

Test for significant difference between conditions using a Wilcoxon rank sum test:

```{r}
print(
  wilcox.test(
    formula=control_flow_inst_prop~condition,
    data=filter(solutions_inst_exec_data),
    exact=FALSE,
    conf.int=TRUE)
)
```

In case you're curious, here's all categories of instructions:

```{r}
melted <- melt(
  solutions_inst_exec_data,
  variable.name = "inst_type",
  value.name = "inst_type_prop",
  measure.vars=c(
    "math_inst_prop",
    "module_inst_prop",
    "memory_inst_prop",
    "regulation_inst_prop",
    "control_flow_inst_prop",
    "thread_inst_prop",
    "task_inst_prop",
    "nop_inst_prop"
  )
)

ggplot( melted, aes(x=inst_type, y=inst_type_prop, color=condition) ) +
  geom_boxplot() +
  scale_color_discrete(
    name="Condition:",
    breaks=c("memory", "both"),
    labels=c("Regulation-disabled", "Regulation-enabled")
  ) +
  xlab("Instruction type") +
  ylab("Proportion of instructions in execution trace") +
  coord_flip() +
  theme(legend.position="bottom")
```

## Visualizing an evolved gene regulatory network

Let's take a closer look at a successful gene regulatory network.

```{r}
trace_id <- 23997
```

Specifically, we'll be looking at the solution evolved in run id `r trace_id` (arbitrarily selected).


### Evolved regulatory network

We use the igraph package to draw this program's gene regulatory network.

```{r, message=FALSE}
# Networks!
graph_nodes_loc <- paste0(working_directory, "data/igraphs/reg_graph_id-", trace_id, "_nodes.csv")
graph_edges_loc <- paste0(working_directory, "data/igraphs/reg_graph_id-", trace_id, "_edges.csv")
graph_nodes_data <- read.csv(graph_nodes_loc, na.strings="NONE")
graph_edges_data <- read.csv(graph_edges_loc, na.strings="NONE")

network <- graph_from_data_frame(
  d=graph_edges_data,
  vertices=graph_nodes_data,
  directed=TRUE
)

# Setup edge styling
E(network)$color[E(network)$type == "promote"] <- "#FCE640"
E(network)$lty[E(network)$type == "promote"] <- 1
E(network)$color[E(network)$type == "repress"] <- "#441152"
E(network)$lty[E(network)$type == "repress"] <- 1

network_out_name <- paste0(working_directory, "imgs/case-study-id-", trace_id, "-network.svg")

draw_network <- function(net, write_out, out_name) {
  if (write_out) {
    svg(out_name, width=4,height=1.5)
    # bottom, left, top, right
    par(mar=c(0.2,0,1,0.5))
  }
  plot(
    net,
    edge.arrow.size=0.4,
    edge.arrow.width=0.75,
    edge.width=2,
    vertex.size=10,
    vertex.label.cex=0.65,
    curved=TRUE,
    vertex.color="grey99",
    vertex.label.color="black",
    vertex.label.family="sans",
    layout=layout.circle(net)
  )
  legend(
    x = "bottomleft",      ## position, also takes x,y coordinates
    legend = c("Promoted", "Repressed"),
    pch = 19,              ## legend symbols see ?points
    col = c("#FCE640", "#441152"),
    bty = "n",
    border="black",
    xpd=TRUE,
    title = "Edges"
  )
  if (write_out) {
    dev.flush()
    dev.off()
  }
}

draw_network(network, TRUE, network_out_name)
draw_network(network, FALSE, "")
```


